---
title: | 
  |  \huge Econometric Methods:
  |  Solutions to Empirical Exercise 5.1
  |  Chapter 5: Regression with a Single Regressor: Hypothesis Tests and Confidence Intervals
  |  Stock \& Watson, 3$^{rd}$ Edition
author: 
- \Large Zaeen de Souza\footnote{}
- \Large  Deepti Goel\footnote{\footnotesize Solution key prepared jointly by Zaeen and Deepti. R code and presentation in Rmarkdown by Zaeen.}
date: | 
  |
  | Azim Premji University
  | `r format(Sys.time(), '%d %B %Y')`
output: pdf_document
latex_engine: pdflatex
fontsize: 11pt
header-includes:
- \usepackage{booktabs}
- \usepackage{amsmath}
- \usepackage{dcolumn}
- \usepackage{lscape}
- \usepackage{xcolor}
- \usepackage{mathpazo}
- \usepackage{domitian}
urlcolor: blue
linkcolor: blue
---

```{r setup, include=FALSE}


# in case these are not installed - uncomment and then run.
# install.packages("readxl")
# install.packages("ggplot2")

# Loading excel files
library(readxl)

# Making graphs
library(ggplot2) 

# Setting working directory
setwd("C:/Users/admin/cse Dropbox/Zaeen de Souza/Files From Deepti/Problem Sets - Latex, Code, Rmd/econometric-methods/Chapters/Chapter 5 Hypo. Testing and CI Single Reg/Dataset")

# Loading the data as 'pset_data'
pset_data <- read_excel("Earnings_and_Height.xlsx")
```
\newpage
\tableofcontents
\newpage

# Background: Empirical Exercise 5.1

These are the solutions to **E5.1** from **Chapter 5** of \textit{Introduction to Econometrics (Updated Third edition)} by Stock \& Watson. You should have the following on your computer in order to check answers/run the code and follow the questions in this assignment:

-   An updated version of `R` and `Rstudio`.
-   The following packages installed: 
    - `ggplot2`
    - `readxl`
    - `stargazer`
-   The dataset called `height_and_earnings`.
-   The data description pdf to understand the variables being used.

# Reading guide
All the code needed to complete the assignments is within this document. R code will be in a grey box and will look like this:
```{r, eval=F, echo=T}
summary(iris)
```

And all R output i.e what R shows you once you run some code, will have `#` signs next to it, and will look like this:

```{r, eval=T, echo=F}
summary(iris$Sepal.Length)
```
As far as possible these guides will show the \textbf{exact output} that comes from running code in `R`, and at times will use formatted tables made in latex. The results themselves, will be identical. Some things to note, that might make output look different accross different computers:

- R reports things like p-values using scientific notation, but some computers report the numbers with many trailing zeros.

- If you have an old version of `R` or `Rstudio` it is highly recommended that you update it using the following code:

```{r, eval=F, echo=TRUE}
# Use this to update R from within RStudio
install.packages("installr")
library(installr)
# This last command, will open up a download prompt; choose yes/no accordingly.
updateR()
```

For updating `Rstudio`, un-install your version of `RStudio`, and download a fresh version from the `RStudio` website.
\newpage

# Loading the data and libraries

The following code sets the working directory to the folder where you have downloaded the data, loads the libraries needed fo the assignment and loads the excel dataset.

```{r, eval=F, echo=TRUE}
# Loading excel files
library(readxl)
# Making graphs
library(ggplot2) 

# Setting working directory - this is unique to your computer
setwd("~Zaeen de Souza/Chapter 5 Hypo. Testing and CI Single Reg")

# Loading the data as 'pset_data'
pset_data <- read_excel("Earnings_and_Height.xlsx")
```

# E5.1 Problem Context
On the text website, www.pearsonglobaleditions.com/Stock_Watson, you will find the data file Earnings_and_Height, which contains data on earnings, height, and other characteristics of a random sample of U.S. workers. A detailed description is given in Earnings_and_Height_Description, also available on the website. In this exercise, you will investigate the relationship between earnings and height.
\newpage

# Exercise E5.1


## A. Run a regression of Earnings on Height.

```{r, eval=TRUE, echo=TRUE}
model_1 <- lm(earnings ~ height, data = pset_data)
summary(model_1)
```

### A.i Is the estimated slope statistically significant?

The slope is statistically significant. Under the null hypothesis of no effect (i.e $\beta_{H_0}=0$), we can calculate the T-statistic for the estimated slope using:

$$
T_{actual} = \frac{\hat\beta_{1}-\beta_{H_0}}{SE(\hat\beta)},\ \ \ \ where, \ \beta_{H_0}=0
$$
This gives a T-statistic of $14.016$ and the p-value is $<0.01$, which is enough information for us to reject $H_{0}$, i.e the null that the estimated slope $\hat\beta_{1}=\beta_{H_0}=0$, and accept $H_{1}$ i.e the alternative hypothesis that $\hat\beta\neq 0$.
\
\
You can manually calculate the t-statistic by

```{r, eval=T, echo=T}
# retrieving the stored coefficient from the saved model object
# dividing it by the saved standard error - 
# Note: it retains the name of the variable (height)
t <- round(model_1$coefficients[2]/coef(summary(model_1))[, "Std. Error"][2],3)
t
```

### A.ii Construct a 95% confidence interval for the slope coefficient.

Given the estimated slope which is $707.67$ and the standard error of $50.49$, we can construct the 95\% confidence interval using the following formula

$$
\hat \beta \pm (1.96 \times SE(\hat\beta))
$$
Where $\hat \beta$ is the estimated slope, $1.96$ is the critical value from the standard normal distribution and $SE(\hat\beta)$ is the standard error of the estimated slope. Taking the values directly from the regression output, we have,
$$
707.67 \pm (1.96 \times 50.49) = [608.71, 806.63]
$$
The `R` code to automate the calculation of the interval is below. You can verify that the hand calculation matches the `R` output by,
```{r, eval = T, echo=TRUE}
confint(model_1)
```

## B. Repeat (a) for women.

```{r, eval=T, echo=T}
model_2 <- lm(earnings ~ height, data = subset(pset_data, sex == 0))
summary(model_2)
```

### B.i Is the estimated slope statistically significant?

The slope is statistically significant---looking at p-value for the t-test which is $<0.01$, we can reject the null that the estimated slope for female workers is equal to 0.

### B. ii Construct a 95% confidence interval for the slope coefficient.
Using the formula from `A.ii`, we have the following 95\% confidence interval for our estimated slope for female workers.
$$
511.2  \pm (1.96 \times 98.9) = [317.36, 705.04]
$$
```{r, eval=T, echo=TRUE}
confint(model_2)
```
## C. Repeat (a) for men.

```{r, eval=T, echo=T}
model_3 <- lm(earnings ~ height, data = subset(pset_data, sex == 1))
summary(model_3)
```

### C.i Is the estimated slope statistically significant?

The slope is statistically significant---looking at p-value for the t-test which is $<0.01$, we can reject the null hypothesis that the estimated slope for male workers is equal to 0.


### C. ii Construct a 95% confidence interval for the slope coefficient.
Using the formula from `A.ii`, we have the following 95\% confidence interval for our estimated slope for male workers,
$$
1306.9 \pm (1.96 \times 100.8) = [1109.33, 1504.47]
$$
```{r, eval=T, echo=TRUE}
confint(model_3)
```
## D. Test the null hypothesis that the effect of height on earnings is the same for men and women. (Hint: See Exercise 5.15.)

Let the difference between the two coefficients be given by,

$$
\hat\beta_{\Delta}=\hat\beta_{m}-\hat\beta_{f}
$$
Where the subscripts $m$ and $f$ denote male and female, respectively. Recall that the standard error of $\hat\beta_{\Delta}$ is given by the following formula.:
$$
SE(\hat\beta_{\Delta})=\sqrt{[SE(\hat\beta_{m})]^{2})+[SE(\hat\beta_{f})]^{2}}
$$
Note as well, some people might find this version easier to work with, which has been slightly re-arranged---both will give you the same answer
$$
SE(\hat\beta_{\Delta})=([SE(\hat\beta_{m})]^{2})+[SE(\hat\beta_{f})]^{2})^\frac{1}{2}
$$
Substituting, values from the previous questions, we have the difference in coefficients given by,

$$
\hat\beta_{\Delta} = (1306.9-511.2) = 795.7
$$
And using the formula to calculate the standard error of the difference, again, using the estimated coefficients, we have the following:
$$
SE(\hat\beta_{\Delta})= \sqrt{[(100.8^{2})+(98.9^{2})]}=101.2
$$
Therefore it follows from the standard formula for the confidence interval, that the 95\% confidence interval within which the difference falls is given by,

$$
795.7 \pm (1.96 \times 101.2)=[597.3,994.1]
$$
\newpage

## E. One explanation for the effect on height on earnings is that some professions require strength, which is correlated with height. Does the effect of height on earnings disappear when the sample is restricted to occupations in which strength is unlikely to be important?

We use the subset function again, to run our regression on a subset of the data for occupations that are likely to not have a major need for strength. Note the use of the `'|'` operator that allows us to use an `'or'` condition within the subset function\footnote{Note as well, that this is just one method of using multiple conditions in the subset function. You could also experiment with functions such as \%in\%}.

```{r, eval=T, echo=TRUE}
model_4 <- lm(earnings ~ height, 
              data = subset(pset_data, 
                            occupation == 1 | 
                            occupation == 2 | 
                            occupation == 4 | 
                            occupation == 5 | 
                            occupation == 6 | 
                            occupation == 8))
summary(model_4)
```

The correlation between height and wages does not disappear when restricting the analysis only to people who work in the following occupations.

- 1 = Exec/Manager
- 2 = Professionals
- 4 = Sales
- 5 = Administration
- 6 = Household service
- 8 = Other Service

In fact, the estimated slope is larger than the coefficient from the full sample regression.