---
title: | 
  |  \huge Econometric Methods:
  |  Solutions to Empirical Exercise 5.3
  |  Chapter 5: Regression with a Single Regressor: Hypothesis Tests and Confidence Intervals
  |  Stock \& Watson, 3$^{rd}$ Edition
author: 
- \Large Zaeen de Souza\footnote{}
- \Large  Deepti Goel\footnote{\footnotesize Solution key prepared jointly by Zaeen and Deepti. R code and presentation in Rmarkdown by Zaeen.}
date: | 
  |
  | Azim Premji University
  | `r format(Sys.time(), '%d %B %Y')`
output: pdf_document 
latex_engine: pdflatex
fontsize: 11pt
header-includes:
- \usepackage{booktabs}
- \usepackage{amsmath}
- \usepackage{dcolumn}
- \usepackage{lscape}
- \usepackage{mathpazo}
- \usepackage{domitian}
- \usepackage{xcolor}
urlcolor: blue
linkcolor: blue
---



```{r setup, include=FALSE}
# in case these are not installed - uncomment and then run.
# install.packages("readxl")
# install.packages("ggplot2")

# Loading excel files
library(readxl)

# Making graphs
library(ggplot2) 

# Setting working directory
setwd("C:/Users/admin/cse Dropbox/Zaeen de Souza/Files From Deepti/Problem Sets - Latex, Code, Rmd/econometric-methods/Chapters/Chapter 5 Hypo. Testing and CI Single Reg/Dataset")
# Loading the data as 'pset_data'
pset_data <- read_excel("birthweight_smoking.xlsx")
```
\newpage
\tableofcontents
\newpage

# Background: Empirical Exercise 5.3

These are the solutions to **E5.3** from **Chapter 5** of \textit{Introduction to Econometrics (Updated Third edition)} by Stock \& Watson. You should have the following on your computer in order to check answers/run the code and follow the questions in this assignment:

-   An updated version of `R` and `Rstudio`.
-   The following packages installed: 
    - `ggplot2`
    - `readxl`
    - `stargazer`
-   The dataset called `birthweight_smoking`.
-   The data description pdf to understand the variables being used.

# Reading guide
All the code needed to complete the assignments is within this document. R code will be in a grey box and will look like this:
```{r, eval=F, echo=T}
summary(iris)
```

And all R output i.e what R shows you once you run some code, will have `#` signs next to it, and will look like this:

```{r, eval=T, echo=F}
summary(iris$Sepal.Length)
```
As far as possible these guides will show the \textbf{exact output} that comes from running code in `R`, and at times will use formatted tables made in latex. The results themselves, will be identical. Some things to note, that might make output look different accross different computers:

- R reports things like p-values using scientific notation, but some computers report the numbers with many trailing zeros.

- If you have an old version of `R` or `Rstudio` it is highly recommended that you update it using the following code:

```{r, eval=F, echo=TRUE}
# Use this to update R from within RStudio
install.packages("installr")
library(installr)
# This last command, will open up a download prompt; choose yes/no accordingly.
updateR()
```

For updating `Rstudio`, un-install your version of `RStudio`, and download a fresh version from the `RStudio` website.
\newpage

# Loading the data and libraries

The following code sets the working directory to the folder where you have downloaded the data, loads the libraries needed fo the assignment and loads the excel dataset.

```{r, eval=F, echo=TRUE}
# Loading excel files
library(readxl)
# Making graphs
library(ggplot2) 

# Setting working directory - this is unique to your computer
setwd("~Zaeen de Souza/Chapter 5 Hypo. Testing and CI Single Reg")

# Loading the data as 'pset_data'
pset_data <- read_excel("birthweight_smoking.xlsx")
```

# E5.3 Problem Context
On the text website, www.pearsonglobaleditions.com/Stock_Watson, you will find the data file Birthweight_Smoking, which contains data for a random sample of babies born in Pennsylvania in 1989. The data include the baby’s birth weight together with various characteristics of the mother, including whether she smoked during the pregnancy. A detailed description is given in Birthweight_Smoking_Description, also available on the website. In this exercise you will investigate the relationship between birth
weight and smoking during pregnancy.
\newpage

# Exercise E5.3

# A. In the sample:

## A.i. What is the average value of Birthweight for all mothers?
The average birthweight for all mothers in the sample is $3382.934$.
```{r, eval=T, echo=TRUE}
mean(pset_data$birthweight, na.rm = T)
```

## A.ii. For mothers who smoke?
The average birthweight for all mothers who do not smoke is $3178.832$.
```{r, eval=T, echo=TRUE}
mean(pset_data$birthweight[pset_data$smoker == 1], na.rm = T)
```

## A.iii. For mothers who do not smoke?
The average birthweight for all mothers who smoke is $3432.06$.
```{r, eval=T, echo=TRUE}
mean(pset_data$birthweight[pset_data$smoker == 0], na.rm = T)
```


# B

## B.i. Use the data in the sample to estimate the difference in average birth weight for smoking and nonsmoking mothers.

The difference in birthweight between these groups is $3432.060-3178.832=253.228$. This version of the t=test syntax essentially tells `R` to run the test using the variable `smoker` as a grouping variable. Note the use of the `~` to indicate 'by group'.

```{r, eval=T, echo=TRUE}
# t-test
test <- t.test(birthweight ~ smoker, data = pset_data, var.equal = T)
test
```
We can see that the p-value is $<0.01$ and as such, we can conclude that the difference in birthweight between these two groups is significant. Note, that here, the binary variable `smoker` takes the value 1 for smokers, and 0 for non-smokers.
\
\
Note, as well, that the mean birthweight in each group in the t-test output, matches exactly with our estimates from section `A.i` and `A.ii`. To replicate the t-test by hand, follow the steps in Problem set 1 in question `B.iii`.


## B.ii. What is the standard error for the estimated difference in (i)?
We can recover the standard error of the estimated difference by accessing the the different items stored in the object we created called `test`. Note, that the the formula for the hand calculation is given in the last two problem sets---it should be straightforward to replicate here too. Here, we show how to extract it from an existing `R` function.
```{r, eval = T, echo=TRUE}
test$stderr
```

## B.iii. Construct a 95% confidence interval for the difference in the average birth weight for smoking and nonsmoking mothers.
The confidence interval is given as part of the t-test output, and is $[200.3831,\ 306.0736]$. To manually calculate it we can:

$$
253.228 \pm (10.96 \times 26.95149) = [200.4031 , 306.0529]
$$
Similar to the standard error, the confidence interval of the estimated difference is also part of the t-test output that `R` provides. We can access it from the stored object using the following:
```{r, eval=T, echo=TRUE}
test$conf.int
```
It also provides the level, chosen by the user in the `level=` argument of the t-test function. The default is 0.95 (95\%). The minor difference in the hand calculation and the `R` output, is due to rounding. 
\newpage

# C. Run a regression of Birthweight on the binary variable Smoker.

We want to estimate the following linear model using OLS:

$$
Birthweight_{i} = \alpha + Smoker_{i} + \varepsilon_{i}
$$
Where smoker takes the value 1 if mother $i$ smoked and 0, if she didn't.
```{r, eval=T, echo=TRUE}
model_1 <- lm(birthweight ~ smoker, data = pset_data)
summary(model_1)
```

## C.i. Explain how the estimated slope and intercept are related to your answers in parts (a) and (b).

The slope coefficient is the difference in mean birth weight between these two groups. The intercept
is the average birth weight of the children of non-smoking mothers. Formally,

$$
Y =
\begin{cases}
\alpha + \varepsilon & \text{if $Smoker = 1$}, \\
\alpha + \beta_{1}Smoker + \varepsilon & \text{if $Smoker = 1$}.
\end{cases}
$$

Taking expectations,

$$
E(Y | Smoker=0) = \alpha
$$

$$
E(Y | Smoker=1) = \alpha + \beta_1
$$
The following is an example of a fully saturated model with one binary variable (Namely, `Smoker`). Therefore, by definition $E(Y | Smoker=1) = \alpha + \beta_{1}Smoker$ and therefore, $E(\varepsilon | Smoker=1)=0$. Intuitively, if we are interested in estimates of $\alpha$, $\beta_1$, we can see that they measure the following quantities:

$$
\alpha = E(Y | Smoker = 0)
$$

Substituting for $\alpha = E(Y | Smoker = 0)$, 
$$
\beta_{1} = E(Y | Smoker = 1) - E(Y | Smoker = 0)
$$

## C.ii. Explain how the $SE(\hat\beta_{1})$ is related to your answer in b(ii).

It is the same. We have already shown (in the above answer) that regression fits a conditional
mean function—when an outcome is regressed on a single binary variable, the ratio of the resulting slope coefficient and its standard error, are the same as the t-test statistic for the difference between mean outcomes for the two groups identified by the binary variable. 


## C.iii. Construct a 95% confidence interval for the effect of smoking on birth weight.

The hand calculation is as follows:

$$
-253.228 \pm (1.96 \times 26.95) = [-200.406, -306.05]
$$
Using code we have, 

```{r, eval=TRUE, echo=TRUE}
confint(model_1)
```
# D. Do you think smoking is uncorrelated with other factors that cause low birth weight? That is, do you think that the regression error term, say ui, has a conditional mean of zero, given Smoking (Xi)? (You will investigate this further in Birthweight and Smoking exercises in later chapters.)

As mentioned earlier, because this is a fully saturated model, it has the conditional mean zero property by definition, and so it follows that smoking is uncorrelated with epsilon. However, that does not imply that the difference in conditional means, i.e.,  the slope coefficient, has a causal interpretation. It cannot be unambiguously interpreted as the causal effect of smoking because smoking is a voluntary decision made by an individual and other lifestyle choices may simultaneously affect the decision to smoke and having a child with low birthweight.

